---
title: "Gender & Age by Education"
description: |
  SEX BY AGE BY EDUCATIONAL ATTAINMENT FOR THE POPULATION 18 YEARS AND OVER <br /> Table: B15001 - Variables used: 4-42, 45-83.
output_dir: "_site"
site: distill::distill_website
---


```{r, message=FALSE,echo=FALSE}
#all_vars_acs5 <- load_variables(year = 2019, dataset = "acs5")
```



```{r Census API, message = FALSE, echo=FALSE}
#census_api_key("55c2d218ad663e521c268658bf0c9c5c301fef30", install = TRUE, overwrite = TRUE)
#readRenviron("~/.Renviron") 
```


```{r packages, message = FALSE}
#Loading the required packages
library(tidycensus)
library(tidyverse)
library(ggbeeswarm)
library(ggridges)
library(viridis)
library(hrbrthemes)
library(leaflet)
library(sf)
library(packcircles)
library(tigris)
options(tigris_use_cache = TRUE)
library(ggiraph)
library(shiny)
```




From the ACS 2019 5-year survey, we looked specifically at the educational attainment of MI residents 18 years and above. Our variables were already segmented by binary gender. <br /> <br />

Below you will find Geographical visualizations that provide information on the following variables: <br /> <br />
-male 18 to 24 with a High School diploma <br />
-male 18 to 24 without a diploma <br />
-female 18 to 24 with a High School diploma <br />
-female 18 to 24 without a diploma <br />
-male 25 to 34 with a High School diploma <br />
-male 25 to 34 without a diploma <br />
-female 25 to 34 with a High School diploma <br />
-female 25 to 34 without a diploma <br />
-male 35 to 44 with a High School diploma <br />
-male 35 to 44 without a diploma <br />
-female 35 to 44 with a High School diploma <br />
-female 35 to 44 without a diploma <br />
-male 45 to 64 with a High School diploma <br />
-male 45 to 64 without a diploma <br />
-female 45 to 64 with a High School diploma <br />
-female 45 to 64 without a diploma <br />
-male Above 65 with a High School diploma <br />
-male Above 65 without a diploma <br />
-female Above 65 with a High School diploma <br />
-female Above 65 without a diploma <br />



```{r defining age variables, message =FALSE, echo=FALSE}
#defining age variables by aggregating them by age, gender, and high school completion
male18to24noHSdiploma <- c("B15001_004","B15001_005")
male18to24HSdiploma <- c("B15001_006","B15001_007","B15001_008","B15001_009","B15001_010")
male25to34noHSdiploma <- c("B15001_012","B15001_013")
male25to34HSdiploma <- c("B15001_014","B15001_015","B15001_016","B15001_017","B15001_018")
male35to44noHSdiploma <- c("B15001_020","B15001_021")
male35to44HSdiploma <- c("B15001_022","B15001_023","B15001_024","B15001_025","B15001_026")
male45to64noHSdiploma <- c("B15001_028","B15001_029")
male45to64HSdiploma <- c("B15001_030","B15001_031","B15001_032","B15001_033","B15001_034")
maleAbove65noHSdiploma <- c("B15001_036","B15001_037")
maleAbove65HSdiploma <- c("B15001_038","B15001_039","B15001_040","B15001_041","B15001_042")
female18to24noHSdiploma <- c("B15001_045","B15001_046")
female18to24HSdiploma <- c("B15001_047","B15001_048","B15001_049","B15001_050","B15001_051")
female25to34noHSdiploma <- c("B15001_053","B15001_054")
female25to34HSdiploma <- c("B15001_055","B15001_056","B15001_057","B15001_058","B15001_059")
female35to44noHSdiploma <- c("B15001_061","B15001_062")
female35to44HSdiploma <- c("B15001_063","B15001_064","B15001_065","B15001_066","B15001_067")
female45to64noHSdiploma <- c("B15001_069","B15001_070")
female45to64HSdiploma <- c("B15001_071","B15001_072","B15001_073","B15001_074","B15001_075")
femaleAbove65noHSdiploma <- c("B15001_077","B15001_078")
femaleAbove65HSdiploma <- c("B15001_079","B15001_080","B15001_081","B15001_082","B15001_083")
```



```{r new dataset, message=FALSE,echo=FALSE}
#r dataset with new age variables
state_age_estimate <- get_acs(
  geography = "county",
  state =  "MI",
  variables = c(male18to24noHSdiploma, male18to24HSdiploma, male25to34noHSdiploma, male25to34HSdiploma,
                male35to44noHSdiploma, male35to44HSdiploma, male45to64noHSdiploma, male45to64HSdiploma,
                maleAbove65noHSdiploma, maleAbove65HSdiploma, female18to24noHSdiploma,
                female18to24HSdiploma, female25to34noHSdiploma, female25to34HSdiploma,
                female35to44noHSdiploma, female35to44HSdiploma, female45to64noHSdiploma,
                female45to64HSdiploma, femaleAbove65noHSdiploma, femaleAbove65HSdiploma),
  survey = "acs5",
  year = 2019,
  output = "wide",
  geometry = FALSE
)
```


```{r rowwise summaries, message=FALSE,echo=FALSE}
#rowwise summaries for age dataset
state_age_summary <- state_age_estimate %>%
  rowwise() %>%
  mutate(
    male18to24noHSdip = sum(B15001_004E,B15001_005E),
    male18to24HSdip = sum(B15001_006E,B15001_007E,B15001_008E,B15001_009E,B15001_010E),
    male25to34noHSdip = sum(B15001_012E,B15001_013E),
    male25to34HSdip = sum(B15001_014E,B15001_015E,B15001_016E,B15001_017E,B15001_018E),
    male35to44noHSdip = sum(B15001_020E,B15001_021E),
    male35to44HSdip = sum(B15001_022E,B15001_023E,B15001_024E,B15001_025E,B15001_026E),
    male45to64noHSdip = sum(B15001_028E,B15001_029E),
    male45to64HSdip = sum(B15001_030E,B15001_031E,B15001_032E,B15001_033E,B15001_034E),
    maleAbove65noHSdip = sum(B15001_036E,B15001_037E),
    maleAbove65HSdip = sum(B15001_038E,B15001_039E,B15001_040E,B15001_041E,B15001_042E),
    female18to24noHSdip = sum(B15001_045E,B15001_046E),
    female18to24HSdip = sum(B15001_047E,B15001_048E,B15001_049E,B15001_050E,B15001_051E),
    female25to34noHSdip = sum(B15001_053E,B15001_054E),
    female25to34HSdip = sum(B15001_055E,B15001_056E,B15001_057E,B15001_058E,B15001_059E),
    female35to44noHSdip = sum(B15001_061E,B15001_062E),
    female35to44HSdip = sum(B15001_063E,B15001_064E,B15001_065E,B15001_066E,B15001_067E),
    female45to64noHSdip = sum(B15001_069E,B15001_070E),
    female45to64HSdip = sum(B15001_071E,B15001_072E,B15001_073E,B15001_074E,B15001_075E),
    femaleAbove65noHSdip = sum(B15001_077E,B15001_078E),
    femaleAbove65HSdip = sum(B15001_079E,B15001_080E,B15001_081E,B15001_082E,B15001_083E)
  ) %>%
  select(NAME, male18to24noHSdip, male18to24HSdip, male25to34noHSdip, male25to34HSdip,
         male35to44noHSdip, male35to44HSdip, male45to64noHSdip, male45to64HSdip, maleAbove65noHSdip,
         maleAbove65HSdip, female18to24noHSdip, female18to24HSdip, female25to34noHSdip, female25to34HSdip,
         female35to44noHSdip, female35to44HSdip, female45to64noHSdip, female45to64HSdip, 
         femaleAbove65noHSdip, femaleAbove65HSdip) %>%
  arrange(NAME) %>%
  ungroup()
```



```{r cleaning the data, message=FALSE,echo=FALSE}
#separating NAME for age data
state_age_summary <- separate(
  state_age_summary,
  NAME,
  into = c("county", "state"),
  sep = ", "
)
#tidy age dataset
MI_age_by_edu <- state_age_summary %>% 
  select(-state)
#age data set with totals
county_totals <- MI_age_by_edu %>%
  rowwise() %>%
  mutate(TOTAL = sum(c(male18to24noHSdip, male18to24HSdip, male25to34noHSdip,
                       male25to34HSdip, male35to44noHSdip, male35to44HSdip,
                       male45to64noHSdip, male45to64HSdip, maleAbove65noHSdip,
                       maleAbove65HSdip, female18to24noHSdip, female18to24HSdip,
                       female25to34noHSdip, female25to34HSdip, female35to44noHSdip,
                       female35to44HSdip, female45to64noHSdip, female45to64HSdip,
                       femaleAbove65noHSdip, femaleAbove65HSdip)))
#Comparing age by gender by educational attainment in MI.
MI_age <- county_totals  %>%
  pivot_longer(!c(county,TOTAL), names_to = "variables", values_to = "count") %>%
  relocate(TOTAL, .after = count)
```



```{r percentages, message=FALSE,echo=FALSE}
#age by gender by education for all counties 
MI_age_percent <- MI_age %>%
  mutate(percent = 100 * (count / TOTAL)) %>%
  select(county, variables, percent)
#Group-wise age by gender by educational (AGE) analysis
largest_AGE_group <- MI_age_percent %>%
  group_by(county) %>%
  filter(percent == max(percent))
  
least_AGE_group <- MI_age_percent %>%
  group_by(county) %>%
  filter(percent == min(percent))
```




In the plot below, each point represents a county in Michigan. The plot shows the distributions of the percentage of people represented by each variable per county in Michigan. <br /> <br /> 

```{r Variable plot, message=FALSE,echo=FALSE}
ggplot(MI_age_percent, aes(x = variables, y = percent, color= percent)) +
  geom_quasirandom(alpha = 0.5) +
  coord_flip() +
  scale_color_viridis_c(guide = "none") +
  labs(title = "Michigan Counties Sex by Age by Educational Attainment for the Population 18 and Over.",
       caption = "Data Source: 2015-2019 ACS\n") +
  theme(plot.title = element_text(hjust = 1, size =12))
```


We decided to create box plots that would further allow us to investigate the noticeable spread on all variables representing people with high school diplomas as shown below. <br /> <br />




```{r most interesting variables, message=FALSE}
cool_age_vars <- MI_age_percent %>%
  group_by(county) %>%
  filter(row_number() %in% c(2,4,6,8,10,12,14,16,18,20))

#box plot
ggplot(cool_age_vars, aes(x = variables, y = percent)) +
  geom_boxplot(
    
    # custom boxes
    color="blue",
    fill="blue",
    alpha=0.5,
    
    # Notch?
    notch=TRUE,
    notchwidth = 0.8,
    
    # custom outliers
    outlier.colour="red",
    outlier.fill="red",
    outlier.size=3
  ) +
  labs(title = "Individuals with a High School Diploma",
       caption = "Data Source: 2015-2019 ACS\n") +
  theme(plot.title = element_text(hjust = 0.5, size =12)) +
  coord_flip()
```

We noticed that the more females than males between 45 and 64 years had high school diplomas as seen from the means in the plot above although the difference is barely noticeable. <br /> <br />
Interestingly, there are significantly more females than males above 65 years who have high school diplomas in Michigan. <br /> <br />

To investigate this further, we decided to create interactive maps to show how the values of these four variables are distributed in all Michigan counties. <br /> <br />


The map below shows the count of all males ages 45 to 65 without High School Diplomas in each county. <br /> <br />

```{r leaflet map 1, message=FALSE, echo=FALSE}
#Male 45 to 64 with no High School Diploma
data1 <- get_acs(geography = "county",
                 variables = c(male45to64noHSdiploma),
                state = "MI",
                geometry = TRUE) %>%
  st_transform(4326)
data2 <- get_acs(geography = "tract",
                 variables = c(male45to64noHSdiploma),
                state = "MI",
                geometry = TRUE) %>%
  st_transform(4326)
bins1 <- c(0,50,150,300,500,1000,2000,3000,4000,5000,6000,7000,8000,9000,10000,
          15000,22000)
pala1 <- colorBin("viridis", data1$estimate, bins = bins1)
leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = data1, stroke = FALSE, smoothFactor = 0.2, 
              color = ~pala1(estimate), 
              #label = ~as.character(estimate), 
              popup = paste(data1$NAME, "<br>",
                            "Value: ", data1$estimate, "<br>"),
              fillOpacity = 0.8, 
              group = "Counties") %>%
  addPolygons(data = data2, stroke = FALSE, smoothFactor = 0.2, 
              color = ~pala1(estimate), 
              popup = paste(data2$NAME, "<br>",
                            "Value: ", data2$estimate, "<br>"),
              #label = ~as.character(estimate), 
              fillOpacity = 0.8, 
              group = "Tracts") %>%
  addLegend("bottomright", pal = pala1, values = data1$estimate, 
            title = "Males 45 to \n
            64 Without HS Diploma") %>%
  addLayersControl(overlayGroups = c("Tracts", "Counties")) %>% 
  hideGroup("Tracts")
```


The map below shows the count of all females ages 45 to 65 without High School Diplomas in each county. <br /> <br />

```{r leaflet map 2, message=FALSE,echo=FALSE}
#Female 45 to 64 with no High School Diploma.
data3 <- get_acs(geography = "county",
                 variables = c(female45to64noHSdiploma),
                state = "MI",
                geometry = TRUE) %>%
  st_transform(4326)
data4 <- get_acs(geography = "tract",
                 variables = c(female45to64noHSdiploma),
                state = "MI",
                geometry = TRUE) %>%
  st_transform(4326)
bins2 <- c(0,50,150,300,500,1000,2000,3000,4000,5000,6000,7000,8000,9000,10000,
          15000,20000,22000)
pala2 <- colorBin("viridis", data3$estimate, bins = bins2)
leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = data3, stroke = FALSE, smoothFactor = 0.2, 
              color = ~pala2(estimate), 
              #label = ~as.character(estimate), 
              popup = paste(data3$NAME, "<br>",
                            "Value: ", data3$estimate, "<br>"),
              fillOpacity = 0.8, 
              group = "Counties") %>%
  addPolygons(data = data4, stroke = FALSE, smoothFactor = 0.2, 
              color = ~pala2(estimate), 
              popup = paste(data4$NAME, "<br>",
                            "Value: ", data4$estimate, "<br>"),
              #label = ~as.character(estimate), 
              fillOpacity = 0.8, 
              group = "Tracts") %>%
  addLegend("bottomright", pal = pala2, values = data3$estimate, 
            title = "Females 45 to \n
            64 Without HS Diploma") %>%
  addLayersControl(overlayGroups = c("Tracts", "Counties")) %>% 
  hideGroup("Tracts")
```




The map below shows the count of all males above 65 with a High School Diplomas in each county. <br /> <br />

```{r leaflet map 3, message=FALSE,echo=FALSE}
#Male above 65 with High School Diploma.
data5 <- get_acs(geography = "county",
                 variables = c(maleAbove65HSdiploma),
                state = "MI",
                geometry = TRUE) %>%
  st_transform(4326)
data6 <- get_acs(geography = "tract",
                 variables = c(maleAbove65HSdiploma),
                state = "MI",
                geometry = TRUE) %>%
  st_transform(4326)
bins3 <- c(0,50,150,300,500,1000,2000,3000,4000,5000,6000,7000,8000,9000,10000,
          15000,20000,22000)
pala3 <- colorBin("viridis", data5$estimate, bins = bins3)
leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = data5, stroke = FALSE, smoothFactor = 0.2, 
              color = ~pala3(estimate), 
              #label = ~as.character(estimate), 
              popup = paste(data5$NAME, "<br>",
                            "Value: ", data5$estimate, "<br>"),
              fillOpacity = 0.8, 
              group = "Counties") %>%
  addPolygons(data = data6, stroke = FALSE, smoothFactor = 0.2, 
              color = ~pala3(estimate), 
              popup = paste(data6$NAME, "<br>",
                            "Value: ", data6$estimate, "<br>"),
              #label = ~as.character(estimate), 
              fillOpacity = 0.8, 
              group = "Tracts") %>%
  addLegend("bottomright", pal = pala3, values = data5$estimate, 
            title = "Male above 65 with HS Diploma") %>%
  addLayersControl(overlayGroups = c("Tracts", "Counties")) %>% 
  hideGroup("Tracts")
```



The map below shows the count of all females above 65 with a High School Diplomas in each county. <br /> <br />

```{r message=FALSE,echo=FALSE}
#Female above 65 with High School Diploma.
data7 <- get_acs(geography = "county",
                 variables = c(femaleAbove65HSdiploma),
                state = "MI",
                geometry = TRUE) %>%
  st_transform(4326)
data8 <- get_acs(geography = "tract",
                 variables = c(femaleAbove65HSdiploma),
                state = "MI",
                geometry = TRUE) %>%
  st_transform(4326)
bins4 <- c(0,50,150,300,500,1000,2000,3000,4000,5000,6000,7000,8000,9000,10000,
          15000,20000,22000)
pala4 <- colorBin("viridis", data7$estimate, bins = bins4)
leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = data7, stroke = FALSE, smoothFactor = 0.2, 
              color = ~pala4(estimate), 
              #label = ~as.character(estimate), 
              popup = paste(data7$NAME, "<br>",
                            "Value: ", data7$estimate, "<br>"),
              fillOpacity = 0.8, 
              group = "Counties") %>%
  addPolygons(data = data4, stroke = FALSE, smoothFactor = 0.2, 
              color = ~pala4(estimate), 
              popup = paste(data8$NAME, "<br>",
                            "Value: ", data8$estimate, "<br>"),
              #label = ~as.character(estimate), 
              fillOpacity = 0.8, 
              group = "Tracts") %>%
  addLegend("bottomright", pal = pala4, values = data7$estimate, 
            title = "Male above 65 with HS Diploma") %>%
  addLayersControl(overlayGroups = c("Tracts", "Counties")) %>% 
  hideGroup("Tracts")
```

In the above maps, we see that the area around Detroit has the largest numbers while the counties in the upper peninsula have the least. <br /> <br />
