---
title: "Language and Employment"
description: |
  EDUCATIONAL ATTAINMENT AND EMPLOYMENT STATUS BY LANGUAGE SPOKEN AT HOME <br /> Table: B16010 - Variables used: 01-53
output_dir: "_site"
site: distill::distill_website
#runtime: shiny
search: TRUE
---

```{r define_variables,echo=FALSE}
edu_emp_lang_total <- ("B16010_001")
# less than high school
edu_emp_lang_less_hs <- ("B16010_002")
edu_emp_lang_less_hs_work <- ("B16010_003")
edu_emp_lang_less_hs_work_eng <- ("B16010_004")
edu_emp_lang_less_hs_work_spanish <- ("B16010_005")
edu_emp_lang_less_hs_work_indoEurop <- ("B16010_006")
edu_emp_lang_less_hs_work_asian <- ("B16010_007")
edu_emp_lang_less_hs_work_other <- ("B16010_008")
edu_emp_lang_less_hs_NOTwork <- ("B16010_009")
edu_emp_lang_less_hs_NOTwork_eng <- ("B16010_010")
edu_emp_lang_less_hs_NOTwork_spanish <- ("B16010_011")
edu_emp_lang_less_hs_NOTwork_indoEurop <- ("B16010_012")
edu_emp_lang_less_hs_NOTwork_asian <- ("B16010_013")
edu_emp_lang_less_hs_NOTwork_other <- ("B16010_014")
# high school graduate
edu_emp_lang_hs <- ("B16010_015")
edu_emp_lang_hs_work <- ("B16010_016")
edu_emp_lang_hs_work_eng <- ("B16010_017")
edu_emp_lang_hs_work_spanish <- ("B16010_018")
edu_emp_lang_hs_work_indoEurop <- ("B16010_019")
edu_emp_lang_hs_work_asian <- ("B16010_020")
edu_emp_lang_hs_work_other <- ("B16010_021")
edu_emp_lang_hs_NOTwork <- ("B16010_022")
edu_emp_lang_hs_NOTwork_eng <- ("B16010_023")
edu_emp_lang_hs_NOTwork_spanish <- ("B16010_024")
edu_emp_lang_hs_NOTwork_indoEurop <- ("B16010_025")
edu_emp_lang_hs_NOTwork_asian <- ("B16010_026")
edu_emp_lang_hs_NOTwork_other <- ("B16010_027")
# some college
edu_emp_lang_college <- ("B16010_028")
edu_emp_lang_college_work <- ("B16010_029")
edu_emp_lang_college_work_eng <- ("B16010_030")
edu_emp_lang_college_work_spanish <- ("B16010_031")
edu_emp_lang_college_work_indoEurop <- ("B16010_032")
edu_emp_lang_college_work_asian <- ("B16010_033")
edu_emp_lang_college_work_other <- ("B16010_034")
edu_emp_lang_college_NOTwork <- ("B16010_035")
edu_emp_lang_college_NOTwork_eng <- ("B16010_036")
edu_emp_lang_college_NOTwork_spanish <- ("B16010_037")
edu_emp_lang_college_NOTwork_indoEurop <- ("B16010_038")
edu_emp_lang_college_NOTwork_asian <- ("B16010_039")
edu_emp_lang_college_NOTwork_other <- ("B16010_040")
# bachelors degree or higher
edu_emp_lang_bach_deg <- ("B16010_041")
edu_emp_lang_bach_deg_work <- ("B16010_042")
edu_emp_lang_bach_deg_work_eng <- ("B16010_043")
edu_emp_lang_bach_deg_work_spanish <- ("B16010_044")
edu_emp_lang_bach_deg_work_indoEurop <- ("B16010_045")
edu_emp_lang_bach_deg_work_asian <- ("B16010_046")
edu_emp_lang_bach_deg_work_other <- ("B16010_047")
edu_emp_lang_bach_deg_NOTwork <- ("B16010_048")
edu_emp_lang_bach_deg_NOTwork_eng <- ("B16010_049")
edu_emp_lang_bach_deg_NOTwork_spanish <- ("B16010_050")
edu_emp_lang_bach_deg_NOTwork_indoEurop <- ("B16010_051")
edu_emp_lang_bach_deg_NOTwork_asian <- ("B16010_052")
edu_emp_lang_bach_deg_NOTwork_other <- ("B16010_053")
```

```{r make_dataset,echo=FALSE}
edu_emp_lang_ds <- get_acs(
  geography = "county",
  state =  "MI",
  variables = c(edu_emp_lang_total),
  survey = "acs5",
  year = 2019,
  output = "wide",
  geometry = FALSE
)
# less than high school
edu_emp_lang_less_hs_ds <- get_acs(
  geography = "county",
  state =  "MI",
  variables = c(edu_emp_lang_less_hs, edu_emp_lang_less_hs_work,
                edu_emp_lang_less_hs_work_eng,
                edu_emp_lang_less_hs_work_spanish,
                edu_emp_lang_less_hs_work_indoEurop,
                edu_emp_lang_less_hs_work_asian,
                edu_emp_lang_less_hs_work_other,
                edu_emp_lang_less_hs_NOTwork,
                edu_emp_lang_less_hs_NOTwork_eng,
                edu_emp_lang_less_hs_NOTwork_spanish,
                edu_emp_lang_less_hs_NOTwork_indoEurop,
                edu_emp_lang_less_hs_NOTwork_asian,
                edu_emp_lang_less_hs_NOTwork_other),
  survey = "acs5",
  year = 2019,
  output = "wide",
  geometry = FALSE
)
# high school
edu_emp_lang_hs_ds <- get_acs(
  geography = "county",
  state =  "MI",
  variables = c(edu_emp_lang_hs, edu_emp_lang_hs_work, 
                edu_emp_lang_hs_work_eng,
                edu_emp_lang_hs_work_spanish,
                edu_emp_lang_hs_work_indoEurop,
                edu_emp_lang_hs_work_asian,
                edu_emp_lang_hs_work_other,
                edu_emp_lang_hs_NOTwork,
                edu_emp_lang_hs_NOTwork_eng,
                edu_emp_lang_hs_NOTwork_spanish,
                edu_emp_lang_hs_NOTwork_indoEurop,
                edu_emp_lang_hs_NOTwork_asian,
                edu_emp_lang_hs_NOTwork_other),
  survey = "acs5",
  year = 2019,
  output = "wide",
  geometry = FALSE
)
# some college
edu_emp_lang_college_ds <- get_acs(
  geography = "county",
  state =  "MI",
  variables = c(edu_emp_lang_college, edu_emp_lang_college_work,
                edu_emp_lang_college_work_eng,
                edu_emp_lang_college_work_spanish,
                edu_emp_lang_college_work_indoEurop,
                edu_emp_lang_college_work_asian,
                edu_emp_lang_college_work_other,
                edu_emp_lang_college_NOTwork,
                edu_emp_lang_college_NOTwork_eng,
                edu_emp_lang_college_NOTwork_spanish,
                edu_emp_lang_college_NOTwork_indoEurop,
                edu_emp_lang_college_NOTwork_asian,
                edu_emp_lang_college_NOTwork_other),
  survey = "acs5",
  year = 2019,
  output = "wide",
  geometry = FALSE
)
# bachelor's degree or higher
edu_emp_lang_bach_deg_ds <- get_acs(
  geography = "county",
  state =  "MI",
  variables = c(edu_emp_lang_bach_deg, edu_emp_lang_bach_deg_work,
                edu_emp_lang_bach_deg_work_eng,
                edu_emp_lang_bach_deg_work_spanish,
                edu_emp_lang_bach_deg_work_indoEurop,
                edu_emp_lang_bach_deg_work_asian,
                edu_emp_lang_bach_deg_work_other,
                edu_emp_lang_bach_deg_NOTwork,
                edu_emp_lang_bach_deg_NOTwork_eng,
                edu_emp_lang_bach_deg_NOTwork_spanish,
                edu_emp_lang_bach_deg_NOTwork_indoEurop,
                edu_emp_lang_bach_deg_NOTwork_asian,
                edu_emp_lang_bach_deg_NOTwork_other),
  survey = "acs5",
  year = 2019,
  output = "wide",
  geometry = FALSE
)
```


```{r combining_variables,echo=FALSE}
# combining high school graduate, some college, and bachelors or higher
edu_emp_lang_grad <- c("B16010_015", "B16010_028", "B16010_041")
edu_emp_lang_grad_work <- c("B16010_016", "B16010_029", "B16010_042")
edu_emp_lang_grad_work_eng <- c("B16010_017", "B16010_030", "B16010_043")
edu_emp_lang_grad_work_spanish <- c("B16010_018", "B16010_031", "B16010_044")
edu_emp_lang_grad_work_indoEurop <- c("B16010_019", "B16010_032", "B16010_045")
edu_emp_lang_grad_work_asian <- c("B16010_020", "B16010_033", "B16010_046")
edu_emp_lang_grad_work_other <- c("B16010_021", "B16010_034", "B16010_047")
edu_emp_lang_grad_NOTwork <- c("B16010_022", "B16010_035", "B16010_048")
edu_emp_lang_grad_NOTwork_eng <- c("B16010_023", "B16010_036", "B16010_049")
edu_emp_lang_grad_NOTwork_spanish <- c("B16010_024", "B16010_037", "B16010_050")
edu_emp_lang_grad_NOTwork_indoEurop <- c("B16010_025", "B16010_038", "B16010_051")
edu_emp_lang_grad_NOTwork_asian <- c("B16010_026", "B16010_039", "B16010_052")
edu_emp_lang_grad_NOTwork_other <- c("B16010_027", "B16010_040", "B16010_053")

# combining languages for those with less than high school diploma
employed_lessHS <- c(lessHS_eng = "B16010_004", 
                     lessHS_spanish = "B16010_005", 
                     lessHS_indoEurop = "B16010_006", 
                     lessHS_asian = "B16010_007",
                     lessHS_other = "B16010_008")

UNemployed_lessHS <- c(lessHS_eng = "B16010_010", 
                       lessHS_spanish = "B16010_011", 
                       lessHS_indoEurop = "B16010_012", 
                       lessHS_asian = "B16010_013",
                       lessHS_other = "B16010_014")

## Making new dataset with combined variables
edu_emp_lang_comb_ds <- get_acs(
  geography = "county",
  state =  "MI",
  variables = c(edu_emp_lang_grad, edu_emp_lang_grad_work,
                edu_emp_lang_grad_work_eng, 
                edu_emp_lang_grad_work_spanish, 
                edu_emp_lang_grad_work_indoEurop, 
                edu_emp_lang_grad_work_asian, 
                edu_emp_lang_grad_work_other, 
                edu_emp_lang_grad_NOTwork, 
                edu_emp_lang_grad_NOTwork_eng, 
                edu_emp_lang_grad_NOTwork_spanish, 
                edu_emp_lang_grad_NOTwork_indoEurop, 
                edu_emp_lang_grad_NOTwork_asian, 
                edu_emp_lang_grad_NOTwork_other),
  survey = "acs5",
  year = 2019,
  output = "wide",
  geometry = FALSE
)

# making row-wise summaries for the combined dataset
edu_emp_lang_comb_summary <- edu_emp_lang_comb_ds %>%
  rowwise() %>%
  mutate(
    edu_emp_lang_grad = sum(B16010_015E, B16010_028E, B16010_041E), 
    edu_emp_lang_grad_work = sum(B16010_016E, B16010_029E, B16010_042E),
    edu_emp_lang_grad_work_eng = sum(B16010_017E, B16010_030E, B16010_043E),
    edu_emp_lang_grad_work_spanish = sum(B16010_018E, B16010_031E, B16010_044E),
    edu_emp_lang_grad_work_indoEurop = sum(B16010_019E, B16010_031E, B16010_045E),
    edu_emp_lang_grad_work_asian = sum(B16010_020E, B16010_033E, B16010_046E),
    edu_emp_lang_grad_work_other = sum(B16010_021E, B16010_034E, B16010_047E),
    edu_emp_lang_grad_NOTwork = sum(B16010_022E, B16010_035E, B16010_048E),
    edu_emp_lang_grad_NOTwork_eng = sum(B16010_023E, B16010_036E, B16010_049E),
    edu_emp_lang_grad_NOTwork_spanish = sum(B16010_024E, B16010_037E, B16010_050E),
    edu_emp_lang_grad_NOTwork_indoEurop = sum(B16010_025E, B16010_038E, B16010_051E),
    edu_emp_lang_grad_NOTwork_asian = sum(B16010_026E, B16010_039E, B16010_052E),
    edu_emp_lang_grad_NOTwork_other = sum(B16010_027E, B16010_040E, B16010_053E)
  ) %>%
  select(NAME, edu_emp_lang_grad, edu_emp_lang_grad_work, edu_emp_lang_grad_work_eng, 
         edu_emp_lang_grad_work_spanish, edu_emp_lang_grad_work_indoEurop, 
         edu_emp_lang_grad_work_asian, edu_emp_lang_grad_work_other, 
         edu_emp_lang_grad_NOTwork, edu_emp_lang_grad_NOTwork_eng, 
         edu_emp_lang_grad_NOTwork_spanish, edu_emp_lang_grad_NOTwork_indoEurop, 
         edu_emp_lang_grad_NOTwork_asian, edu_emp_lang_grad_NOTwork_other) %>%
  arrange(NAME) %>%
  ungroup()

# combining languages for facet plots
employed_HS <- c(HS_eng = edu_emp_lang_grad_work_eng, 
                 HS_spanish = edu_emp_lang_grad_work_spanish, 
                 HS_indoEurop = edu_emp_lang_grad_work_indoEurop, 
                 HS_asian = edu_emp_lang_grad_work_asian,
                 HS_other = edu_emp_lang_grad_work_other)

UNemployed_HS <- c(HS_eng = edu_emp_lang_grad_NOTwork_eng, 
                   HS_spanish = edu_emp_lang_grad_NOTwork_spanish, 
                   HS_indoEurop = edu_emp_lang_grad_work_indoEurop, 
                   HS_asian = edu_emp_lang_grad_work_asian,
                   HS_other = edu_emp_lang_grad_work_other)
```

From the ACS 2019 5-year survey, we compared if people are native English speakers (do they speak a different language at home) and if they are employed by their education level for Kent County. <br /> 
 <br />
 
The facet plots below use variables of:<br /> <br /> 
-Percent employed without a high school diploma <br /> 
-Percent unemployed without a high school diploma <br /> 
-Percent employed with a high school diploma <br /> 
-Percent unemployed with a high school diploma <br />
with all of these being broken down by the language spoken at home (as categorized by the US Census):<br /> 
-An Asian language <br />
-Engish <br />
-An Indo-European language<br />
-Spanish <br />
-Other <br /><br />

Observations from the facet plots: <br /> 
In general, these plots helped us to see what languages are spoken in Grand Rapids and the rest of Kent County. Our biggest populations spoke English and Spanish, likely as one would expect for our area. <br />
These first two plots are just for people without a high school diploma. We can see that most of the people who do not have a high school diploma and are employed tend to speak Spanish at home. Additionally, most of the people without a diploma but are unemployed tend to speak English or Spanish. <br />

```{r making_graphs,echo=FALSE}
## facet plots
# employed, less than HS
kent_tracts <- get_acs(geography = "tract",
                       variables = employed_lessHS, 
                       state = "MI",
                       county = "Kent County",
                       geometry = TRUE,
                       summary_var = "B20004_001") 

kent_tracts <- kent_tracts %>% 
  mutate(pct = 100 * estimate / summary_est)

kent_tracts %>% 
  ggplot(aes(fill = pct)) +
  geom_sf(color = NA) +
  labs(title = 'Percent Employed Without HS Diploma by Language Spoken at Home',
       subtitle = "2019 5-year ACS estimates for Kent County",
       caption='Source: ACS Data Table B29002 via the tidycensus R package',
       x = NULL,
       y = NULL)+
  theme(axis.text = element_blank(),  # no lat/long values
        axis.ticks = element_blank()) + # no lat/long ticks
  facet_wrap(~variable, ncol = 3, nrow = 2) +
  scale_fill_viridis_c()

# unemployed, less than hs
kent_tracts <- get_acs(geography = "tract",
                       variables = UNemployed_lessHS, 
                       state = "MI",
                       county = "Kent County",
                       geometry = TRUE,
                       summary_var = "B20004_001") 

kent_tracts <- kent_tracts %>% 
  mutate(pct = 100 * estimate / summary_est)

kent_tracts %>% 
  ggplot(aes(fill = pct)) +
  geom_sf(color = NA) +
  labs(title = 'Percent Unemployed Without HS Diploma by Language Spoken at Home',
       subtitle = "2019 5-year ACS estimates for Kent County",
       caption='Source: ACS Data Table B29002 via the tidycensus R package',
       x = NULL,
       y = NULL)+
  theme(axis.text = element_blank(),  # no lat/long values
        axis.ticks = element_blank()) + # no lat/long ticks
  facet_wrap(~variable, ncol = 3, nrow = 2) +
  scale_fill_viridis_c()
```

More observations from the facet plots: <br /> 
This next set of plots represent people with at least a high school diploma. It is broken up into sections of (1) people with a high school diploma, (2) people with some college, and (3) people with a bachelor's degree or higher.  <br />
Here it is clear that most of the people who have a high school diploma or higher education in Kent County tend to speak English, no matter their employment status. <br />

```{r graphs_with_HS,echo=FALSE}
# employed, hs or higher
kent_tracts_HS <- get_acs(geography = "tract",
                       variables = employed_HS, 
                       state = "MI",
                       county = "Kent County",
                       geometry = TRUE,
                       summary_var = "B20004_001") 

kent_tracts_HS <- kent_tracts_HS %>% 
  mutate(pct = 100 * estimate / summary_est)

kent_tracts_HS %>% 
  ggplot(aes(fill = pct)) +
  geom_sf(color = NA) +
  labs(title = 'Percent Employed With HS Diploma by Language Spoken at Home',
       subtitle = "2019 5-year ACS estimates for Kent County",
       caption='Source: ACS Data Table B29002 via the tidycensus R package',
       x = NULL,
       y = NULL)+
  theme(axis.text = element_blank(),  # no lat/long values
        axis.ticks = element_blank()) + # no lat/long ticks
  facet_wrap(~variable, ncol = 6, nrow = 3) +
  scale_fill_viridis_c()

# UNemployed, hs or higher
kent_tracts_HS <- get_acs(geography = "tract",
                       variables = UNemployed_HS, 
                       state = "MI",
                       county = "Kent County",
                       geometry = TRUE,
                       summary_var = "B20004_001") 

kent_tracts_HS <- kent_tracts_HS %>% 
  mutate(pct = 100 * estimate / summary_est)

kent_tracts_HS %>% 
  ggplot(aes(fill = pct)) +
  geom_sf(color = NA) +
  labs(title = 'Percent Unemployed With HS Diploma by Language Spoken at Home',
       subtitle = "2019 5-year ACS estimates for Kent County",
       caption='Source: ACS Data Table B29002 via the tidycensus R package',
       x = NULL,
       y = NULL)+
  theme(axis.text = element_blank(),  # no lat/long values
        axis.ticks = element_blank()) + # no lat/long ticks
  facet_wrap(~variable, ncol = 6, nrow = 4) +
  scale_fill_viridis_c()
```