---
title: "Poverty Status by Gender"
description: |
  POVERTY STATUS IN THE PAST 12 MONTHS OF INDIVIDUALS BY SEX BY EDUCATIONAL ATTAINMENT <br /> Table: B17003 - Variables used: 4-23.
output_dir: "_site"
site: distill::distill_website
---

```{r, message=FALSE,echo=FALSE}
#all_vars_acs5 <- load_variables(year = 2019, dataset = "acs5")
```



```{r Census API, message = FALSE, echo=FALSE}
#census_api_key("55c2d218ad663e521c268658bf0c9c5c301fef30", install = TRUE, overwrite = TRUE)
#readRenviron("~/.Renviron") 
```

POVERTY STATUS IN THE PAST 12 MONTHS OF INDIVIDUALS BY SEX BY EDUCATIONAL ATTAINMENT
Table b17003



```{r packages, message = FALSE}
#Loading the required packages
library(tidycensus)
library(tidyverse)
library(ggbeeswarm)
library(ggridges)
library(viridis)
library(hrbrthemes)
library(leaflet)
library(sf)
library(packcircles)
library(tigris)
options(tigris_use_cache = TRUE)
library(ggiraph)
library(shiny)
```




```{r defining poverty variables, message=FALSE, echo=FALSE}
#define poverty variables,
poorMales_noHSdiploma <- c("B17003_004")
poorMales_HSdiploma <- c("B17003_005","B17003_006","B17003_007")
poorFemales_noHSdiploma <- c("B17003_009")
poorFemales_HSdiploma <- c("B17003_010","B17003_011","B17003_012")
NotPoorMales_noHSdiploma <- c("B17003_015")
NotPoorMales_HSdiploma <- c("B17003_016","B17003_017","B17003_018")
NotPoorFemales_noHSdiploma <- c("B17003_020")
NotPoorFemales_HSdiploma <- c("B17003_021","B17003_022","B17003_023")
```


```{r new dataset, message=FALSE, echo=FALSE}
#dataset with the new poverty variables
state_poverty_estimate <- get_acs(
  geography = "county",
  state =  "MI",
  variables = c(poorMales_noHSdiploma, poorMales_HSdiploma, poorFemales_noHSdiploma,
                poorFemales_HSdiploma, NotPoorMales_noHSdiploma, NotPoorMales_HSdiploma,
                NotPoorFemales_noHSdiploma, NotPoorFemales_HSdiploma),
  survey = "acs5",
  year = 2019,
  output = "wide",
  geometry = FALSE
)
```


From the ACS 2019 5-year survey, we looked at the poverty status of Michigan residents based on whether they obtained high school diplomas or not. <br /> <br /> 

Below you will find Geographical visualizations that provide information on the following variables: <br /> <br />
-Poor males with no high school diploma <br /> 
-Poor males with a high school diploma <br />
-Poor females with no high school diploma <br />
-Poor females with a high school diploma <br />
-Not poor males with no high school diploma <br />
-Not poor males with a high school diploma <br />
-Not poor females with no high school diploma <br />
-Not poor females with a high school diploma <br />


```{r, rowwise summaries, message=FALSE, echo=FALSE}
#rowwise summary for poverty dataset
state_poverty_summary <- state_poverty_estimate %>%
  rowwise() %>%
  mutate(poorMales_noHSdip = B17003_004E,
         poorMales_HSdip = sum(B17003_005E,B17003_006E,B17003_007E),
         poorFemales_noHSdip = B17003_009E,
         poorFemales_HSdip = sum(B17003_010E,B17003_011E,B17003_012E),
         NotPoorMales_noHSdip = B17003_015E,
         NotPoorMales_HSdip = sum(B17003_016E,B17003_017E,B17003_018E),
         NotPoorFemales_noHSdip = B17003_020E,
         NotPoorFemales_HSdip = sum(B17003_021E,B17003_022E,B17003_023E)
  ) %>%
  select(NAME, poorMales_noHSdip, poorMales_HSdip, poorFemales_noHSdip,
         poorFemales_HSdip, NotPoorMales_noHSdip, NotPoorMales_HSdip,
         NotPoorFemales_noHSdip, NotPoorFemales_HSdip) %>%
  arrange(NAME) %>%
  ungroup()
```


```{r tidying, message=FALSE,echo=FALSE}
#separating county name and state name
state_poverty_summary <- separate(
  state_poverty_summary,
  NAME,
  into = c("county", "state"),
  sep = ", "
)
# Deleting column with repeated state name
MI_poverty_by_edu <- state_poverty_summary %>% 
  select(-state)
#poverty dataset with totals
county_poverty_totals <- MI_poverty_by_edu %>%
  rowwise() %>%
  mutate(TOTAL = sum(c(poorMales_noHSdip, poorMales_HSdip, poorFemales_noHSdip,
                       poorFemales_HSdip, NotPoorMales_noHSdip, NotPoorMales_HSdip,
                       NotPoorFemales_noHSdip, NotPoorFemales_HSdip)))
#comparing poverty status by sex by educational attainment
MI_poverty <- county_poverty_totals %>%
  pivot_longer(!c(county, TOTAL), names_to = "variables", values_to = "count") %>%
  relocate(TOTAL, .after = count)
```



```{r percentages, message=FALSE,echo=FALSE}
#poverty by sex by educational attainment group percentages per county
MI_poverty_percentage <- MI_poverty %>%
  mutate(percent = 100 * (count/TOTAL)) %>%
  select(county, variables, percent)
#Group wise poverty by gender by education (PGE)analysis
largest_PGE_group <- MI_poverty_percentage %>%
  group_by(county) %>%
  filter(percent == max(percent))
smallest_PGE_group <- MI_poverty_percentage %>%
  group_by(county) %>%
  filter(percent == min(percent))
```



From the plot below, each point represents a county in Michigan. Both males and females with high school diplomas and are not poor have the highest percentages in all counties. Not poor males and females without high school diplomas were the least in all Michigan counties. <br /> <br />
The distribution of either of the genders per poverty and educational attainment category is almost similar. <br /> <br />

```{r plot, message=FALSE,echo=FALSE}
#poverty plot
ggplot(MI_poverty_percentage, aes(x = variables, y = percent, color= percent)) +
  geom_quasirandom(alpha = 0.5) +
  coord_flip() +
  scale_color_viridis_c(guide = "none") +
  labs(title = "Michigan Counties Poverty Status by Sex by Educational Attainment",
       caption = "Data Source: 2015-2019 ACS\n") +
  theme(plot.title = element_text(hjust = -1, size =12))
```

From the plot above each point represents a county in Michigan. Females and males with high school diplomas and are not poor have the highest percentages in all counties. The percentages of poor people with high school diplomas were the least in all counties. <br /> <br />


From the plot below, we can see that the means of each of these variables is almost similar. <br /> <br />

```{r interesting variables, message=FALSE, echo=FALSE}
cool_poverty_vars <- MI_poverty_percentage %>%
  group_by(county) %>%
  filter(row_number() %in% c(2,4,6,8,10,12,14,16,18,20))
ggplot(MI_poverty_percentage, aes(x = variables, y = percent)) +
  geom_boxplot(
    
    # custom boxes
    color="blue",
    fill="blue",
    alpha=0.2,
    
    # Notch?
    notch=TRUE,
    notchwidth = 0.8,
    
    # custom outliers
    outlier.colour="red",
    outlier.fill="red",
    outlier.size=3
  ) +
labs(title = "BoxPlot",
       caption = "Data Source: 2015-2019 ACS\n") +
  theme(plot.title = element_text(hjust = 0.5, size =12)) +
  coord_flip()
```