---
title: "Voting Age"
description: |
  CITIZEN, VOTING-AGE POPULATION BY EDUCATIONAL ATTAINMENT <br /> Table: B29002 - Variables used: 01-08 (not 02)
output_dir: "_site"
site: distill::distill_website
#runtime: shiny
search: TRUE
---

```{r define_variables,echo=FALSE}
voting_age_total <- ("B29002_001")
# voting_age_below_9th <- ("B29002_002")
voting_age_hs <- ("B29002_003")
voting_age_hs_grad <- ("B29002_004")
voting_age_college <- ("B29002_005")
voting_age_assoc <- ("B29002_006")
voting_age_bach <- ("B29002_007")
voting_age_grad <- ("B29002_008")
```

```{r make_dataset,echo=FALSE}
voting_age_ds <- get_acs(
  geography = "state",
  state =  "MI",
  # county = "Kent",
  variables = c(voting_age_hs, voting_age_hs_grad, 
                voting_age_college, voting_age_assoc, voting_age_bach,
                voting_age_grad),
  survey = "acs5",
  year = 2019,
  # output = "wide",
  summary_var=voting_age_total,
  geometry = FALSE
)
```

From the ACS 2019 5-year survey, we analyzed a count of those older than 18 and what their education level is to see the graduation rates for the entire population of Michigan. <br /> 
 <br />

Observations from the bar chart: <br /> 
This bar chart shows that less than 10% of the voting age population does not at least have a high school diploma. Furthermore, the largest percentages of the population that can vote have either a high school diploma or some college education. <br />

```{r making_barchart,echo=FALSE}
voting_age_perc %>%
  ggplot(aes(x=variable, y=percent)) +
  geom_bar(stat="identity", fill="#f68060", alpha=.6, width=.4) +
  coord_flip() +
  xlab("") +
  theme_bw()+
  scale_x_discrete(labels = c('High School, No Diploma','High School Graduate', 
                              'Some College, No Degree', 'Associates Degree', 
                              'Bachelors Degree', 'Graduate Degree or Higher'))+
  labs(title = "Percent of MI Voting-Age Population by Educational Attainment", 
       subtitle = "2019 5-year ACS estimates", 
       y = "Percent", 
       x = "Education Level", 
       caption = "Source: ACS Data Table B29002 via the tidycensus R package") 
```
